---
title: "What is Iron Maiden, Taylor Swift and Pharrell Williams singing? 
        A primer on text mining.^[The data and scripsts posted here: https://github.com/wguillioli]"
author: "Walter Guillioli"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
---

## Introduction 
I love music and data science. Purpose of this paper is to apply main text analytics techniques to explore lyrics from different authors.

We will see things like sentiment analysis, word clouds, topic model and network bigrams.

Purpose is to process lyrics and get idea of what they sign without reading lyrics.

A key companino for this was the book tidy text oby x (add link)


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      out.width = '80%', 
                      fig.align = "center",
                      warning = FALSE,
                      message = FALSE) 

rm(list = ls()) #clean environment

require(tidyverse)
require(readr)
require(tidytext)
require(wordcloud)
require(scales)
require(reshape2)
library(forcats)
library(igraph)
library(ggraph)
require(tm)
require(widyr)
require(topicmodels)

data("stop_words")
afinn <- get_sentiments("afinn")
bing <- get_sentiments("bing")
nrc <- get_sentiments("nrc")

#wd <- "C:\\GitHub\\heart_disease\\"

fldr <- "C:/GitHub/music_lyrics/"


# Custom ggplot theme
cust_theme <- theme_minimal() +
  theme(panel.grid.major = element_line(color = "#e1e1e1",  linetype = "dotted"),
        panel.grid.minor = element_blank(),
        legend.position  = "bottom",
        legend.key       = element_blank())

cbPalette <- c("#E69F00", "#999999", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7",
               "#CC6666", "#9999CC", "#66CC99") #2nd line is extra

blue_palette <- c("#004f79", "#0085ca",  "#66b5df", "#474747","#eeeeee")


```


```{r load_data, include=FALSE}

# get all artists from data folder
artists <- list.dirs(path = paste0(fldr, "data/"), 
                     full.names = FALSE, 
                     recursive = FALSE)

# for each artist get all album names into df with one row per each combination
artists_albums_df <- tibble()

for (i in 1:length(artists)){
  albums <- list.dirs(path = paste0(fldr, "data/", artists[i]), 
                      full.names = FALSE, 
                      recursive = FALSE)
  
  for (j in 1:length(albums)){
    artist_album_row <- c(artists[i], albums[j])
    names(artist_album_row) <- c("artist", "album")
    
    artists_albums_df <- bind_rows(artists_albums_df,
                                   artist_album_row)
  }
}

# for each album, get all the song names and lyrics into a df
lyrics_df <- tibble()

for (i in 1:nrow(artists_albums_df)){
  album_path <- paste0(fldr, "data/",
                       artists_albums_df[i,1], "/", #artist
                       artists_albums_df[i,2]) #album
  
  songs <- list.files(album_path)
  
  for (j in 1:length(songs)){
    song_num <- unlist(str_split(songs[j], pattern = "_"))[1]
    song_name <- unlist(str_split(songs[j], pattern = "_"))[2]
    song_path <- paste0(album_path, "/", songs[j])
    lyrics <- read_file(song_path)
    print(song_path)
    
    song_name <- strsplit(song_name, split = "[.]")[[1]][1]
    
    lyric_row <- c(artists_albums_df[i,1],
                   artists_albums_df[i,2],
                   paste0(artists_albums_df[i,1], ": ", artists_albums_df[i,2]),
                   song_num,
                   song_name,
                   lyrics)
    
    names(lyric_row) <- c("artist", "album", "artist_album", 
                          "song_num", "song", "lyrics")
    
    lyrics_df <- bind_rows(lyrics_df, lyric_row)
  }
}

lyrics_df %>%
  group_by(artist_album) %>%
  tally()

```


## Data and Methods
Lyrics from 4 albums were dowloaded from the azlyrics.com website. They were manually copy pasted to text documents and read into R.
Two albums for Iron Maiden - Number of the Beast (1982) and X-Factor (1995). For Taylor Swift, 1989 from 2014. And Pharrell Williams album x from x.

Data was read from txt into data frame. The techniques applied where: list all of them.

48 songs were analyzed from these albums: Iron Maiden: Number of the Beast	9, 		
Iron Maiden: X Factor	11, Pharrell Williams: GIRL	12, Taylor Swift: 1989	16.



```{r print_songs, include=FALSE}

print(lyrics_df[,c(1,2,5)])
print(lyrics_df[,c(5)])


```



```{r data_prep, include=FALSE}

# tokenize lyrics per sentence
lyrics_sentences <- lyrics_df %>%
  unnest_tokens(song_line, lyrics, token = "lines") %>%
  mutate(artist_album_song = paste0(artist_album, song)) %>%
  group_by(artist_album_song) %>%
  mutate(song_line_number = row_number()) %>%
  ungroup() %>%
  select(-artist_album_song)
    
# tokenize lyrics per word
lyrics_words <- lyrics_sentences %>%
  unnest_tokens(word, song_line) 
  
# remove stop words
lyrics_words <- lyrics_words %>%
  anti_join(stop_words)


```

The lyrics were split into invidual words for analysis and common words (stop words) like a, on, the were removed. Table 2  shows number of words for analysis:

```{r print_words, include=FALSE}

# print # of words per album
lyrics_words %>%
  group_by(artist_album) %>%
  summarise(number_of_songs = max(song_num),
            number_of_words = n())

```

**Table X** \
*Number of songs and words per album*

Album |# Songs |# Words
:-----|:-------|:------
Iron Maiden: Number of the Beast |9 |800
Iron Maiden: X Factor |11 |871
Pharrell Williams: GIRL |12 |1437
Taylor Swift: 1989 |16 |2000

Table 1 shows the names of the songs of each album.

\newpage
**Table X** \
*Songs per album*

Taylor Swift: 1989        |Pharrell Williams: GIRL    |Iron Maiden: Sign of the Cross   |Iron Maiden: Number of the Beast
:-------------------------|:--------------------------|:-------------------------------|:--------------------------------
Welcome to NY |Marilyn Monroe |Sign of the Cross | Invaders
Blank Space         |Brand New      |Lord of the Flies | Children Of the Damned
Style	| unter        |Man on the Ledge |The Prisoner
Out in the Woods    |Gush            |Fortunes of War   |22 Acacia Avenue
All you had to... |Happy |Look for the Truth |Number Of The Beast
Shake it off |Come Get It Bae |The Aftermath |Run to the Hills
I wish you would |Gust of Wind |Judgment of Heaven |Gangland
Bad blood |Lost Queen |Blood on the... |Hallowed be thy Name
Wildest Dreams |Freq |Edge of Darkness |Total Eclipse
How you get the girl |Know who you are |2AM	
This love	|It Girl |The Unbeliever	
I know places |Smile		
Clean |			
Wonderland |			
You arein love |			
New romantics |			



## Results

To analyze text we start with x, then do y and then z.

### Top words used on the lyrics

We simply remove stop words and count and get this. Alternative we could use a more advanced way via tfidf to appendix

This is relatively interesting but not useful since we know some artists use more words than others. So let's do the same but by album.

```{r plot_top_album, echo=FALSE}

# most common words per album
lyrics_words %>% 
  count(artist_album, word, sort = TRUE) %>%
  group_by(artist_album) %>%
  slice_max(n, n = 10, with_ties = FALSE) %>% 
    ungroup() %>%
    mutate(word = reorder(word, n)) %>%
    ggplot(aes(n, word, fill = artist_album)) +
    geom_col(fill = blue_palette[2],
             show.legend = FALSE) +
    facet_wrap(~artist_album, scales = "free", ncol = 2) +
    labs(x = NULL,
         y = NULL) +
  cust_theme

 
```


```{r print_clous, echo=FALSE}

#More informally we can see some wordclouds to get an idea of what they talk. This #just confirms what we saw above in more detail.


# word clouds by album
set.seed(666)
#lyrics_words %>%
#  filter(artist_album == albums_to_parse[1]) %>%
#  count(word) %>%
#  with(wordcloud(word, n, max.words = 100))  

#set.seed(666)
#lyrics_words %>%
#  filter(artist_album == albums_to_parse[2]) %>%
#  count(word) %>%
#  with(wordcloud(word, n, max.words = 100))  

#set.seed(666)
#lyrics_words %>%
#  filter(artist_album == albums_to_parse[3]) %>%
#  count(word) %>%
#  with(wordcloud(word, n, max.words = 100))  

#set.seed(666)
#lyrics_words %>%
#  filter(artist_album == albums_to_parse[4]) %>%
#  count(word) %>%
#  with(wordcloud(word, n, max.words = 100)) 

```




### Sentiment evolution in the album
The top words above give an idea of what they talk about but it's just the start. The next thing is to see if they sing about happy, sad or angry stuff. Probably a bit of everything. 

We start with a simple analysis to see if the sentiment is positive or negative. In the chart below we see the sentiment evolution of the albums as the album progress where each bar represents a line in the song. Also note that number of lines varies and Maiden sings with less words.   


```{r plot_sentiment, echo=FALSE}

# plot sentiment change by album (calculated by line)
albumns_sentiment <- lyrics_words %>%
  inner_join(bing) %>%
  count(artist_album, song_line_number, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(sentiment = positive - negative) %>%
  mutate(pos = sentiment >= 0 )

ggplot(albumns_sentiment, aes(song_line_number, sentiment, fill = pos)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~artist_album, ncol = 2, scales = "free_x") +
  cust_theme

#ggplot(albumns_sentiment, aes(song_line_number, sentiment, fill = pos)) +
 # geom_col(fill = blue_palette[2],
  #         show.legend = FALSE) +
  #facet_wrap(~artist_album, ncol = 2, scales = "free_x") +
  #cust_theme

#ggplot(data=albumns_sentiment, aes(x=song_line_number, y=sentiment, #group=artist_album, color=artist_album)) +
#  geom_line(show.legend = FALSE)+
#  #geom_point(show.legend = FALSE) +
#  theme_minimal() +
#  theme(legend.position="bottom")

#ggplot(albumns_sentiment, aes(song_line_number, sentiment, fill = artist_album)) +
 # geom_line(color = blue_palette[2],
  #         show.legend = FALSE) +
  #facet_wrap(~artist_album, ncol = 2) +
  #cust_theme


```

Several points to highlight from this chart. Iron Maiden tends to be very negative while Taylor Swift is mostly negative with some exceptions. Pharrell Williams on the contrary seems to be a very happy and uplifting singer. We will explore why and more in the next section.  


### Main feelings from each album

Another thing we can do is to see the sentiment for each album. Maiden is motly about fear, sadness and anger. While Pharrell sings about Joy and Trust. While Taylor seems to be happier but sometimes is inundated with sadness and fear. As we will see later this is a diffrent fear than the one Maiden shows.

```{r plot_sent_album, echo=FALSE}

# top sentiments per album
nrc_filtered <- nrc %>%
  filter(sentiment != "positive" & sentiment != "negative")

#lyrics_words %>%
#  inner_join(nrc_filtered, relationship = "many-to-many") %>%
#  count(artist_album, sentiment, sort = TRUE) %>%
#  group_by(artist_album) %>%
#  slice(1:5)

lyrics_words %>% 
  inner_join(nrc_filtered, relationship = "many-to-many") %>%
  count(artist_album, sentiment, sort = TRUE) %>%
  #count(artist_album, word, sort = TRUE) %>%
  group_by(artist_album) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
    ungroup() %>%
    mutate(sentiment = reorder(sentiment, n)) %>%
    ggplot(aes(n, sentiment, fill = artist_album)) +
    geom_col(fill = blue_palette[2],
             show.legend = FALSE) +
    facet_wrap(~artist_album, scales = "free") +
    labs(x = NULL,
         y = NULL) +
  cust_theme

```

### What are they afraid of?

A common feeling across all albums is the sentiment of fear. So next we explore the top fear words of each album. 

```{r plot_fear, echo=FALSE}

nrc_fear <- get_sentiments("nrc") %>% 
  filter(sentiment == "fear")

lyrics_words %>%
  #filter(album == "Number of the Beast") %>%
  inner_join(nrc_fear) %>%
  count(artist_album, word, sort = TRUE) %>%
  group_by(artist_album) %>%
  slice_max(order_by = n, n = 7, with_ties = FALSE) %>%
  #ungroup() %>%
    mutate(word = reorder(word, n)) %>%
  ungroup() %>%  
  ggplot(aes(n, word, fill = artist_album)) +
    geom_col(fill = blue_palette[2],
             show.legend = FALSE) +
    facet_wrap(~artist_album, scales = "free") +
    labs(x = NULL,
         y = NULL) +
  cust_theme

```

So even though they are all fear words, the meaning is different. TAke Iron Maiden is talking about fight, kill, murder, beast, war. Whille Pharrell talks about his feelings and moan. Even though hunter appears it's a love song about hunting for his girl. Taylor swift with bad, hate and mad clearly is talking about a bad romance. 

### Positive and negative words 
Exploring the words further we can take a look at the most positive and negative words per album. 

```{r pos_neg, include=FALSE}

# LEAVE OTHER FOR NOW
lyrics_words %>%
  inner_join(bing) %>%
  count(artist_album, sentiment, word, sort = TRUE) %>%
  ungroup() %>%
  group_by(artist_album, sentiment) %>%
  slice_max(n, n = 7, with_ties = FALSE) %>%
    mutate(word = reorder(word, n)) %>%
    ggplot(aes(n, word, fill = sentiment)) +
    geom_col(show.legend = FALSE, fill = blue_palette[3]) +
    facet_wrap(~artist_album+sentiment, 
               scales = "free",
               ncol = 4) +
    labs(x = "Contribution to sentiment",
         y = NULL) +
    #ggtitle(albums_to_parse[i]) +
    theme_minimal()



```


```{r print_top_words_by_sentiment, echo=FALSE}

# ABOVE OR THIS ONE?

albums_to_parse <- unique(lyrics_words$artist_album)

# function that takes an album and prints top 10 pos/neg words
print_pos_neg_words <- function(i){
  
  lyrics_words_top <- lyrics_words %>%
    filter(artist_album == albums_to_parse[i]) %>%
    inner_join(bing) %>%
    count(word, sentiment, sort = TRUE) %>%
    ungroup()
  
  myplot <- lyrics_words_top %>%
    group_by(sentiment) %>%
    slice_max(n, n = 10, with_ties = FALSE) %>% 
    ungroup() %>%
    mutate(word = reorder(word, n)) %>%
    ggplot(aes(n, word, fill = sentiment)) +
    geom_col(show.legend = FALSE,
             fill = blue_palette[1]) +
    facet_wrap(~sentiment, scales = "free_y") +
    labs(x = "Contribution to sentiment",
         y = NULL) +
    ggtitle(albums_to_parse[i]) +
    theme_minimal()
  
  print(myplot)
  
}

print_pos_neg_words(1) #beast

print_pos_neg_words(3) # pharrel





```


Maiden best is interesting. Free is positive but it's actually from running free and it's someone escaping from the police. Hallowed is also from a song called hallowed by the name from someone on death row. so this can be misleading.

Pharrell, appreciate positivism with woo (part of woo-hoo) and in general talks about happiness. Moan comes from sex so huhh.

Taylor we can see that likes to talk about love - good or bad? and shake is that from shake what? can be from shake the body soo...


### How negative is each song?

A next obvious quesitno is beyond words. How does each song contribute to the feelings shows above. For that we calculate a negative ratio of each song which is basically # of negative words over total words so a higher number means a negative song. results are this.



```{r neg_songs, echo=FALSE}
# plot / calculate most negative songs per album

words_per_song <- lyrics_words %>%
  group_by(artist_album, song) %>%
  summarise(words = n())

negative <- bing %>% 
  filter(sentiment == "negative")

negative_words_per_song <- lyrics_words %>%
  semi_join(negative)

negative_ratio <- negative_words_per_song %>%
  group_by(artist_album, song) %>%
  summarise(negativewords =n()) %>%
  left_join(words_per_song, by = c("artist_album", "song")) %>%
  mutate(ratio = round(negativewords/words, 2)) %>%
  ungroup()

ggplot(negative_ratio, aes(song, ratio, fill = artist_album)) +
  geom_col(show.legend = FALSE, fill = blue_palette[2]) +
  facet_wrap(~artist_album, ncol = 2, scales = "free_y") +
  coord_flip() +
  theme_minimal()

```

Makes sense for Maiden cause Chilre of damned talks about homeless kids and 2am talks about contemplating suicidine.

Not surpriseing Williams is an optimistic fellow.

And Taylow, Shake it off, is misleading cause it's not a negative song. it's more about somebody that likes to go out and party and "shakes off" criticisim from others so could be positive. example fo limitation. next is bad blood, which is about a failed relatioship that went from love to hate so it makes sense. 

### Looking beyond one word

As seen, one word is limited so here we look at combinations of two words that give more context. So what are the top 10 combinations of pairs of words per album?


```{r bigram, echo=FALSE}
# 4 n-grams

lyrics_df$lyrics_clean <- removeWords(lyrics_df$lyrics, stop_words$word)

# try here with clean lyrics vs normal
lyrics_bigrams <- lyrics_df %>%
  unnest_tokens(bigram, lyrics, token = "ngrams", n = 2) %>%
  filter(!is.na(bigram))

#lyrics_bigrams %>%
# count(bigram, sort = TRUE)

bigrams_separated <- lyrics_bigrams %>%
  separate(bigram,c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

#bigrams_filtered %>% 
#  count(word1, word2, sort = TRUE)

bigrams_united <- bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ")

bigrams_tfidf <- bigrams_united %>%
  count(artist_album, bigram) %>%
  bind_tf_idf(bigram, artist_album, n) %>%
  arrange(desc(tf_idf))

bigrams_tfidf %>%
  group_by(artist_album) %>%
  slice_max(tf_idf, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  ggplot(aes(tf_idf, fct_reorder(bigram, tf_idf), fill = artist_album)) +
  geom_col(show.legend = FALSE, fill=blue_palette[1]) +
  facet_wrap(~artist_album, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL) +
  theme_minimal()

```

This gives more insight. for example for xfactor above fortune was positive but it's really talking about war fortunes which is an irony for all the bad stuff that happends after the wars. 

Interesting to see how Williams and Pharrell don't use real words and more "ah ah", "eh eh", "woo hoo". 

### Network graphs of these bigrams

naturally we want to see how these words are interconnected to see about topics they talk about. 


```{r nw, echo=FALSE}
# function that takes album name of album and ngram filter 
# and plots the network graph by album
plot_graph <- function(album_to_plot, num){

  bigram_counts <- bigrams_filtered %>%
    filter(artist_album == album_to_plot) %>%
    count(word1, word2, sort = TRUE)
  
  #bigram_counts
  
  bigram_graph <- bigram_counts %>%
    filter(n > num) %>%
    graph_from_data_frame()
  
 # a <- grid::arrow(type = "closed", length = unit(.15, "inches"))
  
  
  set.seed(97702)
  ggraph(bigram_graph, layout = "fr") +
    geom_edge_link() +
    geom_node_point(color = blue_palette[3], size = 3) +
    geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
    theme_minimal()
  
}

```


```{r plotsnw, echo=FALSE}
plot_graph(albums_to_parse[1], 1) #beast


plot_graph(albums_to_parse[4], 3) #taylor
```
taylor, she talks about all she;s gonna do as is shake. but more context is needed cause paople are gonna hate her. another interesting, is she's asking about stay with someone with bad blood.

for williams we can see girl connected to dirty gilr and hey girl and hey baby so sounds about happy love. he also wants to hug.

Maiden beast, see how more context is given. run is connected to varios words and when reading the lyrics one can see that its about running to the hills (escaping), and run for your lives (escaping). 


### Impact of hegated words

Another importantthing to consider is that happy when doing one word is treated as positive so let's say "i am happy" and "i am not happy" will read happy and that is that. so that is why combining two words give context. so it's intersting to see what words were kind of misclassified with negated words before. so i see what words were c("not", "no", "never", "without") after thse ones and how much they misled the scores.



```{r print_negated, echo=FALSE}
# impact of negated words
negation_words <- c("not", "no", "never", "without")

negated_words <- bigrams_separated %>%
  filter(word1 %in% negation_words) %>%
  inner_join(afinn, by = c(word2 = "word")) %>%
  count(artist_album, word1, word2, value, sort = TRUE)

negated_words %>%
  mutate(contribution = n * value) %>%
  arrange(desc(abs(contribution))) %>%
  head(20) %>%
  mutate(word2 = reorder(word2, contribution)) %>%
  ggplot(aes(n * value, word2, fill = n * value > 0)) +
  geom_col(show.legend = FALSE, fill =  blue_palette[1]) +
  labs(x = "Sentiment value * number of occurrences",
       y = "Words preceded by \"not\"") +
  theme_minimal()


```

so for example, worse its the most negative here. but how about it say i am not worse, that would kind of be positive. 

let's take like, if it's never like then we have a problem.

### A note on topic modeling 

Finally imagine we didn't have the albums. we just knew songs. well there is a way to group them by "topic". so I ovbiously tune the algo to 4 topics. see appendix  




## Conclusions
1. one word analysis can be misleading


## References

## Appendix

\newpage
Appendix 1: TFIDF

```{r print_tfidf, echo=FALSE}
#
# tf-idf

album_words <- lyrics_df %>%
  unnest_tokens(word, lyrics) %>%
  count(artist_album, word, sort = TRUE)

total_words <- album_words %>% 
  group_by(artist_album) %>% 
  summarize(total = sum(n))

album_words <- left_join(album_words, total_words)

album_tf_idf <- album_words %>%
  bind_tf_idf(word, artist_album, n)

#album_tf_idf %>%
#  select(-total) %>%
#  arrange(desc(tf_idf))

album_tf_idf %>%
  group_by(artist_album) %>%
  slice_max(tf_idf, n = 10) %>%
  ungroup() %>%
  ggplot(aes(tf_idf, fct_reorder(word, tf_idf), fill = artist_album)) +
  geom_col(show.legend = FALSE, fill = blue_palette[4]) +
  facet_wrap(~artist_album, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL) + 
  theme_minimal()

```

\newpage
Appendix 2

```{r}
print_pos_neg_words(2)
print_pos_neg_words(4)
```

\newpage
Appendix 3

```{r}
plot_graph(albums_to_parse[2], 1) #xf

plot_graph(albums_to_parse[3], 2)

```

\newpage
Appendix 4
```{r topic, echo=FALSE}
# LDA topic model on albums

word_counts_beast <- lyrics_words %>%
  #filter(artist_album == "Iron Maiden: Number of the Beast") %>%
  count(artist_album, word, sort = TRUE)

beast_dtm <- word_counts_beast %>%
  cast_dtm(artist_album, word, n)

#beast_dtm

lda <- LDA(beast_dtm, k = 4, control = list(seed = 1234))
#lda

topics <- tidy(lda, matrix = "beta")

#topics

top_terms <- topics %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>% 
  ungroup() %>%
  arrange(topic, -beta)

top_terms %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE, fill = blue_palette[1]) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered() +
  theme_minimal()

```

if I compre this charts with above it would be easy to identify the 4 albums with these 4 topics.

